Detecting changes in dispersion in COVID-19 incidence time series using a negative binomial model

### 2023-05-30 noise_covid_v2.Rmd
add figures as PDF's and incorporate

## Specific aim

**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint. Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly increases **.

It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al. 2022). Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a reduction in dispersion of cases, potentially modulated by changes in $k_t$. Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.

@@

```{r}
library(knitr)
rat <- data.frame(Mean = c(5, 10), Ratio = c(2.236068, 3.162278))
kable(rat, align = "cc", caption = "Poisson mean to standard deviation ratio", digits=2)
```

@@

```{r, fig.cap="Average Pothoff-Whittinghill (PW) statistic for each NY county plotted against county population size."}
library(tidyverse)
library(changepoint)
source("code/functions_intro.R")

filename <- "data/processed_dat.Rdata"
load(filename)
rm(filename)

scaled_cases <- map(as.data.frame(t(new_cases_subset)), ~ as.vector(scale(.)))

ch_pts <- map(scaled_cases, ~ cpts(cpt.mean(., class = TRUE, penalty = "MBIC", pen.value = 0, method = "PELT", param.estimates = TRUE)))

segments <- map2(.x = as.data.frame(t(new_cases_subset[which(sapply(ch_pts, length) != 0),])), .y = ch_pts[which(sapply(ch_pts, length) != 0)], ~ get_segments(series = .x, changepoints = .y))

mean_pw <- map(segments, ~ get_mean_pw(.))

# Plot the mean PW statistic v. population size

plot(populations$population[which(sapply(ch_pts, length) != 0)], log(unlist(mean_pw)), xlab="Log Population Size", ylab=" Log Mean PW Statistic", col=2, xlim=c(0, 1000000))
```
Methods designed specifically for count data show a similar dependence on population size. The cpt.meanvar function (changepoint package) is designed to detect changes in mean and variance in count data. The number of changepoints identified for a time series increases with its population size.

```{r, fig.cap="Log of the number of changepoints identified by cpt.meanvar in each NY county plotted against log of county population size."}

ch_pts2 <- map(as.data.frame(t(new_cases_subset)), ~ cpts(cpt.meanvar(., class = TRUE, penalty = "MBIC", pen.value = 30, method = "PELT", test.stat = "Poisson", param.estimates = TRUE, minseglen = 2)))

n_ch_pts2 <- map(ch_pts2, ~length(unlist(.)))

plot(populations$population, log(unlist(n_ch_pts2)), xlab="Log Population Size", ylab=" Log Number of Changepoints", col=2)
```
## Modeling proccess mean

@@

A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009). The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time. Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019).

@@

```{r, fig.cap="Example of empirical increase in dispersion at the hypothesized Thanksgiving 2020 changepoint."}
plot(new_cases_subset[1842,], xaxt="n", ylab="Daily Incidence", xlab="Date") #~ inds
abline(v=30, col=2)
#axis(1, inds, inds, cex.axis = .7)
```

@@

```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (Wald)"}
par(mfrow = c(1,2))
state <- "NY"
source("code/wald_counties.R")
signif_index <- state_index[(which(pvals < 0.05))]
source("code/squared.res_v_squared.fitted.R")
source("code/vis_signific_state.R")

state <- "FL"
source("code/wald_counties.R")
```

```{r, fig.cap = "P-values resulting from testing whether the dispersion parameter is smaller after Thanksgiving in each county in New York (LRT)"}
par(mfrow = c(1,2))
state <- "NY"
source("code/lrt_counties.R")

state <- "FL"
source("code/lrt_counties.R")
```

```{r}
library(tigris)
fl <- counties("Florida")
load("data/W_pvalsFL.Rdata")
p <- round(pvals, digits=1)
p[is.na(p)] <- "na"
fl$p <- p

ggplot(fl) +
  geom_sf(aes(fill = p)) 
```

@@

```{r, fig.cap="Wald simulation p-values."}
source("code/validity_and_power_Wald.R")
```
To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed. 
The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean. As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.

@@

```{r, fig.cap="Wald simulation p-values."}
source("code/validity_and_power_LRT.R")
```
There appears to be slight Type I error inflation in both tests.

```{r}
load("data/pvals_sim_Wald.Rdata")
mean(pvals[which(curve_parms$theta1 == curve_parms$theta2)] < 0.05)
mean(pvals[which(curve_parms$theta1 != curve_parms$theta2)] < 0.05)

load("data/pvals_sim_LRT.Rdata")
mean(pvals[which(curve_parms$theta1==curve_parms$theta2)] < 0.05)
mean(pvals[which(curve_parms$theta1!=curve_parms$theta2)] < 0.05)
```

>>>>

**Using county-level SARS-CoV-2 time series from New York state, we investigate dispersion in case count time series and its relationship to a hypothesized changepoint. Specifically, we estimate and compare dispersion on either side of a putative changepoint in variability (Thanksgiving 2020), in order to test the hypothesis that dispersion significantly increases **. 
It was recently found that $k_t$, the time-varying transmission heterogeneity for COVID-19, decreased over time and was significantly associated with interventions to slow spread in Hong Kong (Adam et al. 2022). Since $k_t$ underlies incidence overdispersion, the method developed here will allow practitioners in epidemiology to explore whether local events result in a reduction in dispersion of cases, potentially modulated by changes in $k_t$. Key unanswered questions remain in developing methodology to recover incidence dispersion estimates and using these estimates to test the effect of gatherings and NPIs.

@@

@@

Methods designed specifically for count data show a similar dependence on population size. The cpt.meanvar function (changepoint package) is designed to detect changes in mean and variance in count data. The number of changepoints identified for a time series increases with its population size.

## Modeling proccess mean

@@

A recently proposed negative binomial regression model for time series of counts also accommodates serial dependence (Davis and Wu 2009). The approach presented here is similar, but we use a linear predictor, $\eta$, that includes a natural spline in time. Natural splines are cubic splines which are linear outside of the boundary knots (Perperoglou et al. 2019). The degrees of freedom for the spline were chosen to be three, as there appeared to be no more than one knot in a time series of 60 days around Thanksgiving. In addition, increasing degrees of freedom above three results in overfitting to the time series and thus potentially insignificant changes in dispersion are more readily detected.

@@

![Figure 1. Empirical increase in dispersion](./figures/empirical_increase.pdf)

@@

@@

To ensure that the hypothesis test has adequate power, a simulation using unequal values of theta on either side of Thanksgiving was repeatedly run in order to ensure that the distribution of resulting p-values is heavily right-skewed. 
The resulting probabilities of rejection are visualized below using one of the Gaussian epidemic curves to represent the process mean. As hoped, the proportion of tests that reject is around 0.05 for those instances where the null is true (probability of the normal variable having 0.05 of the probability density below is 0.05). Additionally, the proportion of tests that reject is high in instances where the alternative hypothesis is true.

@@

{![Figure 2A. (1) Simulated p-values for the Wald test](./figures/pvals_sim_Wald.pdf){width=33%}![(2) Simulated p-values for the Likelihood-Ratio test](./figures/pvals_sim_LRT.pdf){width=33%}!
}

There appears to be slight Type I error inflation in both tests.

![Figure 2B. Mapping the Thanksgiving hypothesis across the US](./figures/map_Wald_US.pdf)

![Figure 3. Scanning for breakpoints using the LRT statistic](./figures/LRT_scan_figure.pdf)
